Toward Pandemic-Scale Ancestral Recombination Graphs of SARS-CoV-2

### 2023-05-12 paper.tex
Tweaks to backbone phylo section

\subsection{Backbone phylogeny}
\label{sec:backbone_phylogeny}
%% The ARG is mostly tree-like
Compared to organisms like humans which recombine every generation,
in SARS-CoV2 recombination is rare, with recombination nodes accounting
for <0.3\% of our inferred ancestry (Table~\ref{tab:args}). As a result,
relationships between strains can often be represented by a
single phylogenetic tree, particularly when looking at
a subset of sampled genomes e.g. restricted by time or limited to close
relatives. We expect the ARG to be particularly tree-like
early in the pandemic, when co-infection was less likely.

A classic tree-based summary of SARS-CoV2 ancestry is provided
by the NextStrain project \citep{Hadfield2018-ef}. The trees
available from Nextstrain are based on small subsamples of the
dataset, and early in the pandemic tend to be restricted to
a sample of strains which are not thought to be recombinants.
For validation purposes, we our compare our ARGs with
a downloaded ``all-data'' NextStrain tree, restricted to the time period
covered by each ARG. To enable this, we ``simplified''~\citep{Kelleher2018-xc}
the sc2ts ARGs to a ``backbone'' containing only samples present
the NextStrain tree. This results in a small set of shared
samples (180 for the wide ARG, 88 for the long ARG), none of which are
designated as Pango recombinant lineages by Nextclade (see subsection
\ref{sec:pango_x_lineages}).

These sc2ts backbone phylogenies do contain small amounts of recombination

\begin{figure} \centering
\includegraphics[width=\textwidth]{figures/cophylogeny_wide.pdf}
\caption{\label{fig:cophylogeny}
Tanglegram comparing the leftmost local tree in the Wide ARG
(sampled to mid-2021)
and an ``all-time'' global NextStrain tree (sampled to January 21, 2023). The
trees are pruned down to the samples present in both the trees. Light grey
lines match the corresponding samples between the two trees; black circles
indicate identical sample partitions between the two trees. Terminal branches
are colour-coded according to the Pango lineage status assigned to the tip
samples; no Pango recombinants are shown, as none are shared between the trees.
The tanglegram was generated using the Neighbor-Net algorithm
\citep{Scornavacca2011-mg} implemented in Dendroscope version 3.8.5
\citep{Huson2012-ys}. See Figure~\ref{fig:cophylogeny_long} for the equivalent
cophylogeny for the Long ARG.}
\end{figure}

Figure~\ref{fig:cophylogeny} shows a co-phylogeny of the 88 samples shared
between the wide ARG and a GISAID global ``all-time'' tree from NextStrain
(downloaded on January 21, 2023).
To reduce the ARG to a single tree, we show
the tree in the middle of the viral genome (covering the genome interval 8393 to 21618)
in the Wide ARG;
however, the five other trees in this
simplified ARG show near-identical patterns. Supplementary
Figure~\ref{fig:cophylogeny_long} shows the same comparison for the Long ARG.
It is clear that the backbone topology of the sc2ts tree shows very close
agreement with the NextStrain tree. The sample genomes cluster by their
assigned Pango lineage status, and many variants and their descendents form
identical monophyletic clades in both the trees (e.g., the Alpha and Delta VoC
clades, labelled).

Figure~\ref{fig:cophylogeny} also reveals some notable differences between the
trees. Firstly, the \texttt{sc2ts} tree is generally less well resolved,
particularly in early 2020 when sampling density was much lower than
later in the pandemic.
Secondly, there are non-identical sample partitions
near the tips (e.g., in the Alpha clade).
Finally, and most noticably, the branch lengths differ.
As discussed in Section~\ref{sec:node_dating}, the dating of nodes
other than samples in \texttt{sc2ts} is currently quite crude, and
it is likely that branches not ``anchored'' by samples will be
poorly calibrated. See [SOMEWHERE ELSE] for a discussion of how
these issues may be alleviated in future versions of \texttt{sc2ts}.

>>>>

\subsection{Backbone phylogeny}
\label{sec:backbone_phylogeny}
%% The ARG is mostly tree-like
Compared to organisms like humans which recombine every generation,
in SARS-CoV2 recombination is rare, with recombination nodes accounting
for $<$0.3\% of our inferred ancestry (Table~\ref{tab:args}). As a result,
relationships between strains can often be represented by a
single phylogenetic tree, particularly when looking at
a subset of sampled genomes e.g. restricted by time or limited to close
relatives. We expect the ARG to be particularly tree-like
early in the pandemic, when co-infection was less likely.
\begin{figure} \centering
\includegraphics[width=\textwidth]{figures/cophylogeny_wide.pdf}
\caption{\label{fig:cophylogeny}
Tanglegram comparing the leftmost local tree in the Wide ARG
(sampled to mid-2021)
and an ``all-time'' global NextStrain tree (sampled to January 21, 2023). The
trees are pruned down to the samples present in both the trees. Light grey
lines match the corresponding samples between the two trees; black circles
indicate identical sample partitions between the two trees. Terminal branches
are colour-coded according to the Pango lineage status assigned to the tip
samples; no Pango recombinants are shown, as none are shared between the trees.
The tanglegram was generated using the Neighbor-Net algorithm
\citep{Scornavacca2011-mg} implemented in Dendroscope version 3.8.5
\citep{Huson2012-ys}. See Figure~\ref{fig:cophylogeny_long} for the equivalent
cophylogeny for the Long ARG.}
\end{figure}

A classic tree-based summary of SARS-CoV2 ancestry is provided
by the NextStrain project \citep{Hadfield2018-ef}. The trees
available from Nextstrain are based on small subsamples of the
dataset, and early in the pandemic tend to be restricted to
a sample of strains which are not thought to be recombinants.
For validation purposes, we our compare our ARGs with
a downloaded NextStrain tree, restricted to the time period
covered by each ARG. To enable this, we ``simplified''~\citep{Kelleher2018-xc}
the \texttt{sc2ts} ARGs to a ``backbone'' containing only samples present
the NextStrain tree. This results in a small set of shared
samples (180 for the Wide ARG, 88 for the Long ARG), none of which are
assigned to Pango recombinant lineages by Nextclade (see Section
\ref{sec:pango_x_lineages}).

Our \texttt{sc2ts} backbone phylogenies do contain small amounts of recombination:
there are 7 recombination nodes in the backbone of the Wide ARG and 8 in the backbone
of the Long ARG. However, the backbones remain extremely tree-like, as these
recombination events involve minor, local topological rearrangements.
All involve recombination between closely related lineages (e.g.\ sibling lineages)
and a number even represent recombination within a single lineage, in which the parent
lineages above a recombination event coalesce with each other before coalescing
with any other lineages in the phylogeny. These could either represent true
recombination events between closely related strains, which have previously gone unnoticed,
or could be artefacts of the current \texttt{sc2ts} method and its parameterization.

Figure~\ref{fig:cophylogeny} compares the backbone phylogeny of the Wide ARG with
a GISAID global ``all-time'' tree from NextStrain (downloaded on January 21, 2023).
We illustrate the backbone phylogeny by visualising
a single tree in the middle of the viral genome
(covering the interval 8393--21618), since other regions of the 
genome show almost identical topologies.
It is clear that the backbone topology of the \texttt{sc2ts} tree shows very close
agreement with the NextStrain tree. The sample genomes cluster by their
assigned Pango lineage status, and many variants and their descendents form
identical monophyletic clades in both the trees (e.g., the Alpha and Delta VoC
clades, labelled).
Supplementary Figure~\ref{fig:cophylogeny_long} shows the same comparison for the Long ARG, with similar results.

Figure~\ref{fig:cophylogeny} also reveals some notable differences between the
trees. Firstly, the \texttt{sc2ts} tree is generally less well resolved,
particularly in early 2020 when sampling density was much lower than
later in the pandemic.
Secondly, there are non-identical sample partitions
near the tips (e.g., in the Alpha clade).
Finally, and most noticeably, the branch lengths differ.
As discussed in Section~\ref{sec:node_dating}, the dating of nodes
other than samples in \texttt{sc2ts} is currently quite crude, and
it is likely that branches not ``anchored'' by samples will be
poorly calibrated. See [SOMEWHERE ELSE] for a discussion of how
these issues may be alleviated in future versions of \texttt{sc2ts}.
